Synthesizing Quotient Lenses

### 2017-11-03 papers/BijectiveLensSynth/ConferencePaper/old/rebuttal.txt

- Converting a regular expression to DNF may incur an exponential explosion
(as it is akin to determinizing an NFM). Have you encountered examples where
that is the case? Please state this shortcoming.

We have not encountered any examples in practice where this is a problem.
However, we include in our benchmarks cust/cap-prob and cust/2-cap-prob to
highlight this shortcoming.  Furthermore, we discuss the shortcoming in
Speed of Synthesis in the evaluation section.

- Have you considered using a SAT/SMT solver to search the space of
  permutations?

Because of our use of orderings, we are able to find the permutations in
O(n * log(n)) time.  It does not seem likely that a SAT solver would help
here.

@@

      cron_lens = [CRON_DICT <=> CRON {}]

==========================
Review E:
==========================

- Why can’t the user provide regular expressions in first place that 
are aligned?

Describing a single format using a regular expression is a standard
task that programmers understand.  Aligning two regular expressions is
considerably more difficult.  Due to the presence of translations,
figuring out what needs to be aligned with is no so easy.
Consider the two regular expressions (A|B)C* and
(A|C(C*)A)|(B|B(C*)C).  We believe it is quite nontrivial for a user
to identify which equivalences must be applied to reach two
well aligned lenses.  Furthermore, the aligned regular expressions
could become difficult to read and maintain: (A|B)(e+C+CC* C) is less
understandable than (A|B)C* .  Surely, avoiding this complexity is a
win for the user.  Fortunately, we are successful in our efforts to do
alignment automatically so we can relieve programmers of this burden.

### 2017-11-08 papers/QuotientLensSynth/main.tex

In summary, our main contributions are:
\begin{enumerate}
\item We introduce a novel language of {\em Quotient Regular Expressions}
(QREs) that provide compact, convenient notation for a natural and useful
class of equivalence relations on regular languages (Section~\ref{QRE}).
\item We define a language of {\em QRE lenses}, a subset of bijective quotient
lenses whose types are given by QREs. Our main technical contribution in this
paper is a proof that each QRE lenses $q$ defines the same transformation as a
QRE lens $q'$ which is in normal form. (Section~\ref{QRE-lenses}).
\item Using this normal form, we reduce the problem of {\em synthesizing}
QRE lenses from their QRE types and a set of examples to the previously studied
problem of synthesizing bijective lenses between regular languages 
(Section~\ref{synth}).
\item We extend the Boomerang {\em implementation} with QREs, QRE lens operators,
and QRE lens synthesis and demonstrate its applicability by using it to
synthesize QRE lenses between several real-world data formats from the
{\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

@@

Consider a bidirectional transformation which converts citation records such as
\begin{verbatim}
@Book {conway,
  Author = "J. H. Conway",
  Title = {Regular Algebra and Finite Machines}
}
\end{verbatim}

@@

We wish for this transformation to have the same behaviour on records which
differ only in ``inessential'' details. First, the transformation
should allow whitespace characters to occur arbitrarily between the words in the
label, author and title fields as well as in between them. Second, an
author's name can have the surname followed by a comma and then the other
names separated by arbitrary whitespace, or just all the names separated by
arbitrary whitespace (for example, the names ``Conway, J. H.'' and ``J. H.
Conway'' are equivalent). Lastly, the transformation should also ignore the
order in which the author and title fields occur in each citation record.

@@

\begin{figure}[t]
\begin{lstlisting}
let wsp = [ \n\t\r]
let wsp_to_space   = project wsp$^+$ $\to$ " "
let space_name     = name | (name . wsp_to_space . (name . wsp_to_space)$^*$ . name)
let comma_name   = name | (name . "," . wsp_to_space . (name . wsp_to_space)$^*$ . name)
let space_to_comma = synth comma_name $\Leftrightarrow$ space_name using {("Conway, J. H.", "J. H. Conway")} 
let valid_name = squash comma_name $\to$ space_name using (get space_to_comma)
let wsp_to_and = project (wsp$^+$ . "and" . wsp$^+$) $\to$ " and " 
let bib_names  = valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)$^*$ . valid_name) 
let bib_author = "author = \"" . bib_names . "\""
let title     = word | (word . wsp_to_space . (word . wsp_to_space)$^*$ . word) 
let bib_title = "title = {" . title . "}" 
let bib_perm  = perm (label, bib_author, bib_title) with (project ("," .wsp$^+$) $\to$ ",\n")
let bibtex    = "@book{" . bib_perm . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
  \label{fig:example-qre}
\end{figure}

@@

One potential pitfall with QRE lenses is that the combinators that we define on
them may not preserve this normal form. For example, the composition $q'; q$ of
QRE lenses $q$ and $q'$ might not be equivalent to any quotient lens that has
the canonical normal form of a bijective lens with canonizers at the edges.
Our main technical contribution in this paper is a proof that each QRE lenses
$q$ defines the same transformation as a QRE lens $q'$ which is in normal form.

@@

\noindent However, while \cd{q_1 ; q_2} is not in normal form, our main theorem
shows that there is another QRE lens \cd{q : bibtex} $\Leftrightarrow$ \cd{xml} of the same type as
\cd{q_1 ; q_2} such that \cd{q_1 ; q_2} and \cd{q} define the same
transformation. This proof will allow justify our approach for synthesizing
quotient lenses so that the programmer can \saz{replace all of the code in
  Figure~\ref{fig:example-lens} with a call to the synthesis prodedure:}

@@

\subsection{Syntax of Quotient Regular Expressions}
The language of Quotient Regular Expressions (QREs) is given by the following
grammar:  \saz{Use the same font that we use above---I added a \textbackslash kw
macro}
\begin{align*}
q := \; &R \sep \project{R}{s} \sep \squash{q}{q'}{f} \sep
\perm{q_1, \ldots, q_n}{q} \\
  & | \; \normalize{R}{R'}{f} \sep q \semicolon q' \sep q \cdot q' \sep (q \sep q') \sep q^*,
\end{align*}
where $R$ ranges over regular expressions, $f$ ranges over functions between
regular languages, and $s$ ranges over character strings.

\saz{We use the notation $\mathcal{L}(R)$ to mean the language accepted by the
  regular expression $R$.}

\subsection{Semantics of Quotient Regular Expression}
In the \bibtex{} to EndNote example, we encountered the $\project{R}{s}$,

\noindent $\squash{q}{q'}{f}$ and $\perm{q_1, \ldots, q_n}{q}$ QREs, as well as
the regular operators applied to these, as in

@@

One combinator that we haven't yet introduced is sequential composition.
The $q \semicolon q'$ QRE successively applis
two equivalence relations to a regular language. For example, in the \bibtex{}
to EndNote example, when defining the \cd{valid_name} QRE which maps
\cd{comma_name} into \cd{space_name}, we first defined the equivalence
relations on the \cd{comma_name} and \cd{space_name} representations before
defining the \cd{valid_name} QRE. Using the composition combinator, we could
instead have defined \cd{valid_name} as follows:
\saz{Instead of replaying the \cd{valid_name} example, perhaps use a simpler
  one---even an artificial example might be better.}

\begin{lstlisting}
(* Define a QRE which matches sends all comma names to a space name equivalent *)
let wsp = [ \n\t\r] 
let whole_space_name  = name | (name . wsp+ . (name . wsp+)* . name) 
let whole_comma_name = name | (name . "," . wsp+ . (name . wsp+)* . name) 
let whole_space_to_comma = 
synth whole_comma_name <=> whole_space_name using {("Conway, J. H.", "J. H. Conway")} 
let whole_valid_name = 
squash whole_comma_name -> whole_space_name using (get whole_space_to_comma)

(* Define QRE that sends space names to their canonical representation *)
let wsp_to_space = project wsp+ $\to$ " "
let space_name  = name | (name . wsp_to_space . (name . wsp_to_space)* . name)

(* Compose the two QREs *)
let valid_name = whole_valid_name ; space_name
\end{lstlisting}

@@

The equivalence relations definable by QREs are a strict subset of those that
can be expressed in Boomerang. For instance, in Boomerang, one can map a
regular language described by $R$ to a regular language described by $S$, with
$\mathcal{L}(S) \not \subseteq \mathcal{L}(R)$, whereas each QRE $q$ canonizes
a regular language described by $R$ to regular subset $\mathcal{L}(R')
\subseteq \mathcal{L}(R)$ described by $R'$.

The net effect of these design decisions means that each QRE $q$ enables us to
simultaneously express
\begin{enumerate}
  \item a regular expression $W(q)$ (the ``whole'' of $q$),
  \item an equivalence relation $\eqrel{q}$ on $\mathcal{L}(W(q))$,
  \item a regular expression $K(q)$ (the ``kernel'' of $q$)
  such that $\mathcal{L}(K(q))$ forms a complete set of representatives for
  $\eqrel{q}$, and
  \item a ``canonizing'' function $\canonizer(q):\mathcal{L}(W(q))
  \longrightarrow \mathcal{L}(K(q))$ which given any $w \in \mathcal{L}(W(q))$,
  computes $\canonizer(q)(w)$ as the unique $k$ in $\mathcal{L}(K(q))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{q}$
  \end{enumerate}
  Observe that since $\canonizer(q)$ computes $\canonizer(q)(w)$ as the unique
  $k$ in $\mathcal{L}(K(q))$ such that $w$ is equivalent to $k$ mod $\eqrel{q}$,
  then the equivalence classes of $\eqrel{q}$ are the same as the fibres of
  $\canonizer(q)$.

\subsection{QREs and ambiguity}
Before giving the formal semantics for QREs, we address the issue of {\em
ambiguity} which comes up when defining QREs. If $q$ and  $q'$ are QREs, then
when applying the regular combinators to QREs, we require that if a string $s$
matches any of the regular expressions $W(q) \cdot W(q')$, $W(q) \sep W(q')$,
$W(q)^*$, $K(q) \cdot K(q')$, $K(q) \sep K(q')$, $K(q)^*$, then $s$ matches
that regular expression in only one way. This unambiguity condition is called
{\em strong unambiguity}, and is necessary, firstly because it ensures that the
canonizing function of a QRE is well-defined.

@@

\saz{Is this where we should explain the rules of  Figure~\ref{fig:qrerules}?}

\subsection{Semantics of QREs}

\saz{This needs a bit more setting up / explanation. Explain what the reader
  should expect to see in the figures.}
The semantics for QREs that formally define the whole language $W(q)$, the
kernel language $K(q)$, the equivalence relation $\eqrel{q}$ and the
canonizing function $\canonizer(q)$ are given in
Figures~\ref{fig:wk},~\ref{fig:relations} and~\ref{fig:canonizers}.
\begin{figure}[t]
\saz{composition notation}
  \centering
  \[
    \begin{array}{l@{\quad}l@{\quad}l}

      q & W(q) & K(q) \\ \hline
      R & R & R \\
      R \mapsto s & W(q) & s \\
      \squash{R}{R'}{f} & R \sep R' & R' \\
      \normalize{R}{R'}{f} & R & R' \\
      q_1 \circ  q_2 & W(q_2) & K(q_1) \\
      q_1 \cdot q_2 & W(q_1) \cdot W(q_2) & K(q_1) \cdot K(q_2) \\
      q_1 \sep q_2 & W(q_1) \sep W(q_2) & K(q_1) \sep K(q_2) \\
      q^* & W(q)^* & K(q)^* \\
    \end{array}
  \]
\[
\begin{array}{r@{\quad}l}
W( \perm{q_1, \ldots, q_n}{q} ) = &
\bigcup \limits_{\sigma \in S_n} W(q_{\sigma(1)}) \cdot W(q) \cdot \ldots \cdot
W(q) \cdot W(q_{\sigma(n)})
\\
K( \perm{q_1, \ldots, q_n}{q} ) = &
 K(q_1) \cdot K(q) \cdot \ldots \cdot K(q) \cdot K(q_n) 
\end{array}
\]
  \caption{Whole and Kernel Regular Expressions\bcp{try putting the 
      last two clauses into the same format as the
      others (using $\mathit{qq} = ...$ and maybe reducing spacing around
      cdots, etc.)}}
  \label{fig:wk}
\end{figure}

\begin{figure}[t]
  \centering
\saz{Fix the projection case}
\[
    \begin{array}{l@{\quad}l@{\quad}l} 
      w \; \equiv_R \; w' &\iff& w = w' \\
      w \; \equiv_{R \mapsto s} \; w' \text{ for all }w, w'\\
      w \; \equiv_{\squash{R}{R'}{f}} \; w' &\iff& f(w) = w'
      \text{ or } w = w' \\
      w \; \equiv_{\normalize{R}{R'}{f}} \; w' &\iff&
      f(w)=f(w') \text{ or }w = w'\\
      w \; \equiv_{q_2 \circ q_1} \; w' &\iff& \exists k, k' \in
  \mathcal{L}(K(q_2)) \text{ such that } w \; \equiv_{q_1} \; k, \; w' \;
  \equiv_{q_1} \; k', \text{ and } k \; \equiv_{q_2} \; k'\\
      w \; \equiv_{q_1 \cdot q_2} \; w'  &\iff& w = r_1
      \cdot r_2, \; w' = {r'}_1 \cdot {r'}_2 \text{ with } r_1 \; \equiv_{q_1}
      \; {r'}_1, \; r_2 \; \equiv{q_n} \; {r'}_2\\
      w \; \equiv_{q' \sep q} \; w' &\iff& w \; \equiv_{q_1} \; w'
      \text{ or } \; w \; \equiv_{q_2} \; w'\\
      w \; \equiv_{q*} \; w' &\iff& w = r_1 \cdot \ldots \cdot r_n, \; w'
      = {r'}_1 \cdot \ldots \cdot {r'}_n \text{ and } r_i \equiv_{q} \; {r'}_i
      \\
      w \; \equiv_{\perm{q_1, \ldots, q_n}{q}} \; w' &\iff& w = r_{\sigma(1)}
      \cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}, \;
    w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
    \cdot {r'}_{\theta(n)} \\
    & & \text{ for some } \sigma, \theta \in S_n, \text{ with } r_i \;
    \equiv_{q_i} \; r'_i \text{ and } s_k \; \equiv_{q} \; s'_{k}
    \end{array}
    \]
  \caption{QRE Equivalence Relations}
  \label{fig:relations}
\end{figure}
\begin{figure}[t]
  \begin{center}
\[
    \begin{array}{l@\quad l @\quad l} 
      \canonizer(R) &=& id_{\mathcal{L}(R)} \\
      \canonizer(R \mapsto s)(w) &=& s\\
      \canonizer(\squash{R}{R'}{f})(w) &=& 
\begin{cases}
f(w) & \text{if } w \in \mathcal{L}(R)\\
w & \text{otherwise}
\end{cases}\\
      \canonizer(\normalize{R}{R'}{f}) &=& f\\
      \canonizer(q' \circ q) &=& \canonizer(q') \circ \canonizer(q)\\
      \canonizer(q' \cdot q) &=& \canonizer(q') \cdot \canonizer(q)\\
      \canonizer(q_1 \sep q_2)(w) &=& 
\begin{cases}
\canonizer(q_1)(w) & \text{if } w \in \mathcal{L}(W(q_1))\\
\canonizer(q_2)(w) & \text{if } w \in \mathcal{L}(W(q_2))\\
\end{cases}\\
      \canonizer(q^*) &=& \canonizer(q)^* \\
    \end{array}
    \]
    \end{center}
    $\canonizer(\perm{q_1, \ldots, q_n}{q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \newline
= (\canonizer(q_1)(r_1)) \cdot (\canonizer(q)(s_1)) \cdot \ldots \cdot
(\canonizer(q_n)(r_n))$
  \caption{QRE canonizers}
  \label{fig:canonizers}
\end{figure}

\begin{figure}[p!]
\centering
\begin{prooftree}
\AxiomC{$R$ is strongly unambiguous}
\UnaryInfC{$\mathit{id}(R)$ is well formed}
\end{prooftree}

\begin{prooftree}
\AxiomC{$s \in W(q)$}
\UnaryInfC{$R \mapsto s$ is well formed}
\end{prooftree}

\begin{prooftree}
\AxiomC{$\mathcal{L}(R) \cap \mathcal{L}(R') = \varnothing$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\BinaryInfC{$squash(R, R', \ell)$ is well formed}
\end{prooftree}

\begin{prooftree}
\AxiomC{${\substack{\forall \sigma \neq \theta, \; W(q_{\sigma(1)}) \cdot W(q)
\cdot \ldots \cdot W(q_{\sigma(n)})\\ \cap W(q_{\theta(1)}) \cdot W(q)
\cdot \ldots \cdot W(q_{\theta(n)}) =\varnothing}}$}
\AxiomC{$q_i, q$ are well formed}
\AxiomC{$\forall \sigma, \; K(q_{\sigma(1)}) \cdot^! K(q) \cdot^! \ldots 
\cdot^! K(q_{\sigma(n)})$}
\TrinaryInfC{$\perm{q_1, \ldots, q_n}{q}$ is well formed}
\end{prooftree}

\begin{prooftree}
\AxiomC{$\mathcal{L}(R') \subseteq \mathcal{L}(R)$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\AxiomC{$f$ is surjective}
\AxiomC{$f = f^2$}
\QuaternaryInfC{$normalize \; (R,R', f)$ is well formed}
\end{prooftree}

\begin{prooftree}
\AxiomC{$q$ is well formed}
\AxiomC{$q'$ is well formed}
\AxiomC{$K(q) = W(q')$}
\TrinaryInfC{$q' \circ q$ is well formed}
\end{prooftree}

\begin{prooftree}
\AxiomC{$q$ is well formed}
\AxiomC{$q'$ is well formed}
\AxiomC{$W(q) \cdot^! W(q')$}
\AxiomC{$K(q) \cdot^! K(q')$}
\QuaternaryInfC{$q \cdot q$ is well formed}
\end{prooftree}

\begin{prooftree}
\AxiomC{$q$ is well formed}
\AxiomC{$q'$ is well formed}
\AxiomC{$W(q) \cap W(q') = \varnothing$}
\TrinaryInfC{$q \cdot q$ is well formed}
\end{prooftree}

>>>>

In summary, our main contributions are:
\begin{enumerate}
\item We introduce a novel language of {\em Quotient Regular Expressions}
(QREs) that provide compact, convenient notation for a natural and useful
class of equivalence relations on regular languages (Section~\ref{QRE}).
\item We define a language of {\em QRE lenses}, a subset of bijective quotient
lenses whose types are given by QREs. Our main technical contribution in this
paper is a proof that each QRE lens $q$ defines the same transformation as a
QRE lens $q'$ that is in normal form. (Section~\ref{QRE-lenses}).
\item Using this normal form, we reduce the problem of {\em synthesizing}
QRE lenses from their QRE types and a set of examples to the previously studied
problem of synthesizing bijective lenses between regular languages 
(Section~\ref{synth}).
\item We extend the Boomerang {\em implementation} with QREs, QRE lens operators,
and QRE lens synthesis and demonstrate its applicability by using it to
synthesize QRE lenses between several real-world data formats from the
{\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

@@

Consider a bidirectional transformation that converts citation records such as
\begin{verbatim}
@Book {conway,
  Author = "J. H. Conway",
  Title = {Regular Algebra and Finite Machines}
}
\end{verbatim}

@@

We wish for this transformation to have the same behaviour on records that
differ only in ``inessential'' details. First, the transformation
should allow whitespace characters to occur arbitrarily between the words in the
label, author and title fields as well as in between them. Second, an
author's name can have the surname followed by a comma and then the other
names separated by arbitrary whitespace, or just all the names separated by
arbitrary whitespace (for example, the names ``Conway, J. H.'' and ``J. H.
Conway'' are equivalent). Lastly, the transformation should also ignore the
order in which the author and title fields occur in each citation record.

@@

\begin{figure}[t]
\begin{lstlisting}
let wsp = [ \n\t\r]

(* project QRE maps arbitrary occurrences of whitespace to a single space *)
let wsp_to_space   = project wsp$^+$ $\to$ " "

(* define name representations with a space and with a comma *)
let space_name     = name | (name . wsp_to_space . (name . wsp_to_space)$^*$ . name)
let comma_name   = name | (name . "," . wsp_to_space . (name . wsp_to_space)$^*$ . name)

(* synthesize a lens that maps the comma representation to the space representation *)
let space_to_comma = synth comma_name $\Leftrightarrow$ space_name using {("Conway, J. H.", "J. H. Conway")}

(* squash QRE maps comma_name to space_name *)
let valid_name = squash comma_name $\to$ space_name using (get space_to_comma)

(* define rest of bibtex fields *)
let wsp_to_and = project (wsp$^+$ . "and" . wsp$^+$) $\to$ " and " 
let bib_names  = valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)$^*$ . valid_name) 
let bib_author = "author = \"" . bib_names . "\""
let title     = word | (word . wsp_to_space . (word . wsp_to_space)$^*$ . word) 
let bib_title = "title = {" . title . "}"

(* perm QRE allows permutations of fields with arbitrary whitespace in between *)
let bib_perm  = perm (label, bib_author, bib_title) with (project ("," .wsp$^+$) $\to$ ",\n")
let bibtex    = "@book{" . bib_perm . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
  \label{fig:example-qre}
\end{figure}

@@

One potential pitfall with QRE lenses is that the combinators that we define on
them may not preserve this normal form. For example, the composition $q'; q$ of
QRE lenses $q$ and $q'$ might not be equivalent to any quotient lens that has
the canonical normal form of a bijective lens with canonizers at the edges.
Our main technical contribution in this paper is a proof that each QRE lenses
$q$ defines the same transformation as a QRE lens $q'$ that is in normal form.

@@

\noindent However, while \cd{q_1 ; q_2} is not in normal form, our main theorem
shows that there is another QRE lens \cd{q : bibtex} $\Leftrightarrow$ \cd{xml}
of the same type as \cd{q_1 ; q_2} such that \cd{q_1 ; q_2} and \cd{q} define
the same transformation. This proof will allow justify our approach for
synthesizing quotient lenses so that the programmer can replace all of the
code in Figure~\ref{fig:example-lens} with a call to the synthesis prodedure:

@@

\subsection{Syntax of Quotient Regular Expressions}
The language of Quotient Regular Expressions (QREs) is given by the following
grammar: 
\begin{align*}
c := \; &R \sep \project{R}{s} \sep \squash{c}{c'}{f} \sep
\perm{c_1, \ldots, c_n}{c} \\
  & | \; \normalize{R}{R'}{f} \sep c \semicolon c' \sep c \cdot c' \sep (c \sep
  c') \sep c^*,
\end{align*}
where $c$ ranges over QREs, $R$ ranges over regular expressions, $f$ ranges over
functions between regular languages, and $s$ ranges over character strings. In
subsequent sections, we will use the the notation $\mathcal{L}(R)$ to mean the
language accepted by the regular expression $R$.

\subsection{Overview of Quotient Regular Expression}
In the \bibtex{} to EndNote example, we encountered the $\project{R}{s}$,

\noindent $\squash{c}{c'}{f}$ and $\perm{c_1, \ldots, c_n}{c}$ QREs, as well as
the regular operators applied to these, as in

@@

One combinator that we haven't yet introduced is sequential composition.
The $c \semicolon c'$ QRE successively applies
two equivalence relations to a regular language. For example, suppose that
we wish to express a name as

\noindent $\kw{last\_name \cdot ``,\text{''} \cdot wsp^+ \cdot first\_name}$,
or as $\kw{first\_name\cdot wsp^+ \cdot last\_name}$, with

\noindent $\kw{first\_name\cdot wsp^+ \cdot last\_name}$ as the canonical
representation. We also wish to project away the extra whitespace in between
\kw{first\_name} and \kw{last\_name} once this identification has been made.
Then we can express this equivalence relation by a composition of a
\textit{squash} QRE and a QRE that projects away the extra whitespace in between
\kw{first\_name} and \kw{last\_name}:

\begin{lstlisting}
(* Regular expressions matching different representations *)
let comma_name = last_name . "," . wsp+ . first_name
let space_name = first_name . wsp+ . last_name

(* Squash comma representation to space representation *)
let comma_to_space = 
synth comma_name <=> space_name using {("Conway, J. H.", "J. H. Conway")} 
let comma_to_space = squash comma_name $\to$ space_name using (get comma_to_space)

(* project whitespace in space_name representation *)
let project_in_space = first_name . (project wsp+ $\to$ " ") . last_name

(* Compose the two QREs *)
let valid_name = project_in_space ; comma_to_space
\end{lstlisting}

@@

The equivalence relations definable by QREs are a strict subset of those that
can be expressed in Boomerang. For instance, in Boomerang, one can map a
regular language described by $R$ to a regular language described by $S$, with
$\mathcal{L}(S) \not \subseteq \mathcal{L}(R)$, whereas each QRE $c$ canonizes
a regular language described by $R$ to regular subset $\mathcal{L}(R')
\subseteq \mathcal{L}(R)$ described by $R'$.

The net effect of these design decisions means that each QRE $c$ enables us to
simultaneously express
\begin{enumerate}
  \item a regular expression $W(c)$ (the ``whole'' of $c$),
  \item an equivalence relation $\eqrel{c}$ on $\mathcal{L}(W(c))$,
  \item a regular expression $K(c)$ (the ``kernel'' of $c$)
  such that $\mathcal{L}(K(c))$ forms a complete set of representatives for
  $\eqrel{c}$, and
  \item a ``canonizing'' function $\canonizer(c):\mathcal{L}(W(c))
  \longrightarrow \mathcal{L}(K(c))$ which given any $w \in \mathcal{L}(W(c))$,
  computes $\canonizer(c)(w)$ as the unique $k$ in $\mathcal{L}(K(c))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{c}$
  \end{enumerate}
  Observe that since $\canonizer(c)$ computes $\canonizer(c)(w)$ as the unique
  $k$ in $\mathcal{L}(K(c))$ such that $w$ is equivalent to $k$ mod $\eqrel{c}$,
  then the equivalence classes of $\eqrel{c}$ are the same as the fibres of
  $\canonizer(c)$.

\subsection{Well Formed QREs and ambiguity}
Before giving the formal semantics for QREs, we address the issue of {\em
ambiguity} which comes up when defining QREs. If $c$ and  $c'$ are QREs, then
when applying the regular combinators to QREs, we require that if a string $s$
matches any of the regular expressions $W(c) \cdot W(c')$, $W(c) \sep W(c')$,
$W(c)^*$, $K(c) \cdot K(c')$, $K(c) \sep K(c')$, $K(c)^*$, then $s$ matches
that regular expression in only one way. This unambiguity condition is called
{\em strong unambiguity}, and is necessary, firstly because it ensures that the
canonizing function of a QRE is well-defined.

@@

\begin{figure}[p!]
\begin{center}
\AxiomC{$R$ is strongly unambiguous}
\UnaryInfC{$\mathit{id}(R)$ \textbf{WF}}
\DisplayProof
\hskip 1.5em
\AxiomC{$s \in \mathcal{L}(R)$}
\UnaryInfC{$\project{R}{s}$ \textbf{WF}}
\DisplayProof
\hskip 1.5em 
\end{center}

\begin{prooftree}
\AxiomC{$c_i, c$ are well formed}
\AxiomC{${\substack{\forall \sigma \neq \theta, \; W(c_{\sigma(1)}) \cdot W(c)
\cdot \ldots \cdot W(c_{\sigma(n)})\\ \cap W(c_{\theta(1)}) \cdot W(c)
\cdot \ldots \cdot W(c_{\theta(n)}) =\varnothing}}$}
\AxiomC{$K(c_1) \cdot^! K(c) \cdot^! \ldots \cdot^! K(c) \cdot^! K(c_n)$}
\TrinaryInfC{$\perm{c_1, \ldots, c_n}{c}$ \textbf{WF}}
\end{prooftree}

\begin{center}
\AxiomC{$c$ \textbf{WF}}
\AxiomC{$c'$ \textbf{WF}}
\AxiomC{$\mathcal{L}(W(c)) \cap \mathcal{L}(W(c')) = \varnothing$}
\AxiomC{$f : \mathcal{L}(W(c)) \longrightarrow \mathcal{L}(W(c'))$}
\QuaternaryInfC{$\squash{c}{c'}{f}$ \textbf{WF}}
\DisplayProof
\end{center}
\begin{center}

\begin{prooftree}
\AxiomC{$\mathcal{L}(R') \subseteq \mathcal{L}(R)$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\AxiomC{$f$ is surjective}
\AxiomC{$f = f^2$}
\QuaternaryInfC{$\normalize{R}{R'}{f}$ \textbf{WF}}
\end{prooftree}

\end{center}
\begin{center}
\AxiomC{$c$ \textbf{WF}}
\AxiomC{$c'$ \textbf{WF}}
\AxiomC{$K(c) = W(c')$}
\TrinaryInfC{$c' \; ; \; c$ \textbf{WF}}
\DisplayProof
\hskip 1.5em
\AxiomC{$c$ \textbf{WF}}
\AxiomC{$W(c)^{*!}$}
\AxiomC{$K(c)^{*!}$}
\TrinaryInfC{$c^*$ \textbf{WF}}
\DisplayProof
\end{center}
\begin{center}

\begin{center}
\AxiomC{$c$ \textbf{WF}}
\AxiomC{$c'$ \textbf{WF}}
\AxiomC{$W(c) \cdot^! W(c')$}
\AxiomC{$K(c) \cdot^! K(c')$}
\QuaternaryInfC{$c \cdot c'$ \textbf{WF}}
\DisplayProof
\end{center}

\begin{center}
\AxiomC{$c$ \textbf{WF}}
\AxiomC{$c'$ \textbf{WF}}
\AxiomC{$W(c) \cap W(c') = \varnothing$}
\TrinaryInfC{$c \cdot c$ \textbf{WF}}
\DisplayProof
\hskip 1.5em
\AxiomC{$c$ \textbf{WF}}
\AxiomC{$W(c)^{*!}$}
\AxiomC{$K(c)^{*!}$}
\TrinaryInfC{$c^*$ \textbf{WF}}
\DisplayProof
\end{center}
\end{center}
  \caption{Well-formed QREs 
}
  \label{fig:qrerules}
\end{figure}

Figure~\ref{fig:qrerules} gives the inferrence rules for derving well-formed
QREs. As expected, the unambiguity conditions come into play when defining QREs
using the regular combinators. For example, the inferrence rule for
concatenation is as follows:
\begin{prooftree}
\AxiomC{$c$ \textbf{WF}}
\AxiomC{$c'$ \textbf{WF}}
\AxiomC{$W(c) \cdot^! W(c')$}
\AxiomC{$K(c) \cdot^! K(c')$}
\QuaternaryInfC{$c \cdot c'$ \textbf{WF}}
\end{prooftree}

\noindent This rule says that we can concatenate two QREs $c$ and $c'$ only when
the concatentions of $W(c)$ and $W(c')$, and $K(c)$ and $K(c')$ are ambiguous.

The most complicated inferrence rule is the rule for the \textit{permutation}
combinator:

\begin{prooftree}
\AxiomC{$c_i, c$ are well formed}
\AxiomC{${\substack{\forall \sigma \neq \theta, \; W(c_{\sigma(1)}) \cdot W(c)
\cdot \ldots \cdot W(c_{\sigma(n)})\\ \cap W(c_{\theta(1)}) \cdot W(c)
\cdot \ldots \cdot W(c_{\theta(n)}) =\varnothing}}$}
\AxiomC{$K(c_1) \cdot^! K(c) \cdot^! \ldots \cdot^! K(c) \cdot^! K(c_n)$}
\TrinaryInfC{$\perm{c_1, \ldots, c_n}{c}$ \textbf{WF}}
\end{prooftree}

\noindent The second hypothesis for the \textit{permutation} rule says that for
any two different permutations $\sigma$ and $\theta$, the languages
$W(c_{\sigma(1)}) \cdot W(c) \cdot \ldots \cdot W(c_{\sigma(n)})$ and
$W(c_{\theta(1)}) \cdot W(c) \cdot \ldots \cdot W(c_{\theta(n)})$ must be
disjoint. This is because any string that matches the \textit{permutation}
combinator matches the language $W(c_{\theta(1)}) \cdot W(c) \cdot \ldots \cdot
W(c_{\theta(n)})$ for some permutation $\theta$, so we require that these
regular expressions be disjoint.

The third hypothesis says that the regular expression $K(c_1) \cdot K(c)
\cdot \ldots \cdot K(c) \cdot K(c_n)$ must be unambiguous. This is because the
underlying lens that maps to or from a \textit{permutation} QRE operates on
the language $K(c_1) \cdot K(c) \cdot \ldots \cdot K(c) \cdot K(c_n)$ which is
the kernel of the \textit{permutation} QRE, and as we mentioned, the underlying
lens will require that the source and target regular expressions of the lens be
unambiguous so that the lens can match strings uniquely.

The rest of the inferrence rules for deriving QREs are given in
Figure~\ref{fig:qrerules}.
\subsection{Semantics of QREs}

As we mentioned before, each QRE $c$ enables us to express:
\begin{enumerate}
  \item a regular expression $W(c)$ (the ``whole'' of $c$),
  \item an equivalence relation $\eqrel{c}$ on $\mathcal{L}(W(c))$,
  \item a regular expression $K(c)$ (the ``kernel'' of $c$)
  such that $\mathcal{L}(K(c))$ forms a complete set of representatives for
  $\eqrel{c}$, and
  \item a ``canonizing'' function $\canonizer(c):\mathcal{L}(W(c))
  \longrightarrow \mathcal{L}(K(c))$ which given any $w \in \mathcal{L}(W(c))$,
  computes $\canonizer(c)(w)$ as the unique $k$ in $\mathcal{L}(K(c))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{c}$
  \end{enumerate}
Figures~\ref{fig:wk} gives the inductive rules that define the whole language
$W(c)$ of a QRE. The two interesting rules are the rules for the
\textit{squash}, \textit{normalize} and \textit{permutation} combinators. If $c
= \squash{c}{c'}{f}$, then the whole language of $c$ is $W(c) \sep W(c')$. This
is because the \textit{squash} combinator takes the whole language $W(c)$ of
$c$, the whole language $W(c')$ of $c'$ and a function $f : \mathcal{L}(W(c))
\longrightarrow \mathcal{L}(W(c'))$, maps $\mathcal{L}(W(c))$ into
$\mathcal{L}(W(c'))$ using $f$ and then canonizes $W(c')$ into $K(c')$ using
$\canonizer(c')$.

For the $\perm{(c_1, \ldots, c_n)}{c}$ combinator, the whole language is the
union of languages of the form $W(c_{\sigma(1)}) \cdot W(c) \cdot \ldots \cdot
W(c) \cdot W(c_{\sigma(n)})$. This is because the \textit{permutation}
combinator allows for the string to match any permutation the $c_i$'s while also
taking into account the separator $c$ in between each of the $c_i$'s. The kernel
of the \textit{permutation} combinator is the language $K(c_1) \cdot K(c) \cdot
\ldots \cdot K(c) \cdot K(c_n)$, because the canonical permutation is the
identity permuation, and then each of the parts of the string that match $c_i$
and $c$ are canonized into $K(c_i)$ and $K(c)$ respectively.

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonizer{}$ function for each QRE. Again, the the least obvious definition
is
